---
title: "EDA Report"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  markdown:
    wrap: 72
---

------------------------------------------------------------------------

# Austin Water Safety: Exploratory Data Analysis

## Names and EIDs:

-   Amanda Lee, awl646

-   Nabil Yusufali, nay277

# Introduction

This analysis will look at water quality data from the City of Austin's
Open Data Portal. It is collected to assess water quality conditions in
the natural creeks, aquifers, and lakes in Austin and surrounding areas.
Citizens have a right to know more about their water quality. Especially
given that we, as students, use the same water everyday. Therefore,
learning more about the water quality could give actionable insights
into how to make better water decisions.

![A picture of Barton Springs, one of Austin's significant groundwater
aquifers. Taken from
<https://www.austintexas.org/listings/barton-springs-pool/4687/>.](https://assets.simpleviewinc.com/simpleview/image/upload/c_limit,q_75,w_1200/v1/crm/austin/011-421ace03bc29117_421acf68-dde1-a643-f8d4f8148337c3cd.jpg)

In an article from Inside Climate News, author Martha Pskowski writes
that "there are 224 public water supply wells that are considered
contaminated", according to the Texas Groundwater Protection Committee's
annual report. For a growing state with less and less surface water
sources, this issue is one that cannot be ignored if Texas hopes to
continue to attract new residents. Water quality testing extends to
agriculture as well. As the "largest consumer of groundwater statewide",
the agricultural sector's handling of water quality impacts the diet of
every Texan.

The data set to be analyzed can be found here:
<https://data.austintexas.gov/Environment/Water-Quality-Sampling-Data/5tye-7ray/about_data>.

Each row is a water quality sample by parameter, date, and location in
Austin, TX. We are primarily interested in variables: data_ref_no,
watershed, site_type, site_name, sample_date, lat_dd_wgs84,
lon_dd_wgs84, param_type, parameter, result, method, medium, and unit to
understand the distribution of water quality across Austin. The
following research questions are investigated in this report:

-   **How do bacteria/pathogen concentrations vary by time of day and
    type of water source? (Nabil)**

    -   It is hypothesized that concentrations will spike during working
        hours (8 AM to 5 PM) since this is most likely when technicians
        are out in the field and collecting samples. Highest
        concentrations will likely be found in sources closer to the
        surface and/or external environment. They are more exposed to
        the elements and thus, bacteria have a higher chance of
        infecting the water.

-   **How do bacteria/pathogen concentrations vary by site type in terms
    of safeness? (Amanda)**

    -   It is hypothesized that bacteria/pathogen concentrations will
        more densely populated towards the city's center in comparison
        to the outer regions. This is because urban areas, with higher
        population densities and more human activities, are more likely
        to contribute more to water contamination. Factors such as urban
        runoff, sewage overflows, and industrial discharges can
        significantly impact water quality in these areas. Additionally,
        a majority of water samples will be measured as safe for human
        interaction.

------------------------------------------------------------------------

# Method

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = TRUE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

library(tidyverse)
library(rpart)
library(rpart.plot)
library(plotROC)
```

```{r}
# Importing 2024 Austin water quality sampling data set 
water <- read.csv('https://raw.githubusercontent.com/Amandawlee/sds322_project/refs/heads/main/Water_Quality_Sampling_Data.csv')

# Data tidying/wrangling
water_tidy <- water |>
  # Rename columns to lowercase
  rename_all(~ str_to_lower(.)) |>
  # Select all relevant columns
  select(data_ref_no, watershed, site_type, site_name, sample_date, lat_dd_wgs84, lon_dd_wgs84, param_type, parameter, result, method, medium, unit) |>
  # Remove any missing NA values with selected columns
  na.omit() |>
  #Filter out since only contained in one observation
  filter(parameter != 'HETEROTROPHIC PLATE COUNT') |>
  # Filter out coordinates that are not in Austin
  filter(lon_dd_wgs84 != 0 | lat_dd_wgs84 != 0) |>
  mutate(date = mdy_hms(sample_date), year = year(date))

head(water_tidy)
```

There are 1,475,965 rows and 24 columns in the original data set. Much
of this data is unconcerned with dangerous chemicals or pathogens and
thus we have limited the data set to bacteria/pathogen samples through
the City of Austin open data portal since this has one of the most
hazardous effects on health. The data set 'water' contains 34,893 rows
and 24 columns.

Additionally, the data was tidied/wrangled with the following changes or
omissions. The resulting data set 'water_tidy' contains 32,758 rows and
15 columns. This means that 2,135 NA rows were dropped from the dataset.
The sample_date variable was converted into two new columns for
R-formatted date and year. After tidying the data, each variable has its
own column, each observation has its own row, and each value has its own
cell.

1.  All column names were converted to lowercase.

2.  Only 15 relevant columns out of 24 were selected for analysis
    (data_ref_no, watershed, site_type, site_name, sample_date,
    lat_dd_wgs84, lon_dd_wgs84, param_type, parameter, result, method,
    medium, unit).

3.  Any missing NA values within the selected columns were removed.

4.  The parameter 'HETEROTROPHIC PLATE COUNT' was filtered out since
    there was only one observation.

5.  Any observations with a latitude or longitude coordinate of 0° will
    be excluded, as these coordinates do not fall within the Austin
    area.

------------------------------------------------------------------------

# Results

## Bacteria/Pathogen Concentration Changes Over Time Per Source and Day

To begin to see the variations in bacteria/pathogen concentrations
between ground and surface water concentrations, we must first
investigate the distributions of each variable involved. We'll first
look at the counts of the types of water sources.

```{r}
water_tidy |>
  ggplot() + 
  #Make bar plot of water sources w/o legend
  geom_bar(aes(y = medium, fill = medium), show.legend = F) +
  labs(
    x = 'Number of Sources',
    y = 'Type of Source',
    title = 'Distribution of Water Sources',
    caption = 'Figure 1: Distribution of water sources in Austin, TX'
  ) + 
  theme_bw()

water_tidy |> #Calculate summary statistics (counts) for water sources
  group_by(medium) |>
  summarize(count = n(),
            percent = 100*n()/nrow(water_tidy))
```

From Figure 1 and the summary statistics, the three largest sources are
1,704 storm water sources, 9,161 ground water sources (\~28% of all
water sources), and 22,531 surface water sources (\~69% of all water
sources). Moving forward, storm water sources will not be included for
analysis since they have wildly different bacteria concentrations and
treatment of storm water varies significantly from other sources
(<https://torrentlab.com/what-is-the-purpose-of-stormwater-testing-and-when-is-it-required/>).

Next, we'll look at the distribution of bacteria concentration across
*all* water sources. This will include the following bacteria: total
coliform bacteria, fecal streptococci, fecal coliform bacteria, E coli
bacteria, and enterococcus. Total coliform bacteria is filtered out
since it only accounts for a single row and has one recorded instance in
2012. The EPA recommends that for a domestic water supply that is to be
treated, concentrations for the previously mentioned bacteria should be
less than 2000 colonies per 100 milliliters. Therefore, the
visualization only looks at bacteria concentrations in excess of 2000
colonies/100 mL.

```{r}
water_tidy |>
  filter(result >= 2000 & result < 10000) |> #Only look at concentrations above 2000 colonies/100 mL and less than 10000 to see general trend
  ggplot() +
  geom_histogram(aes(x = result, fill = parameter), show.legend = F, binwidth = 1000, center = 500) +
  facet_wrap(~parameter) +
  labs(
    x = 'Concentration of Bacteria (Colonies/100 mL)',
    y = 'Frequency',
    title = 'Distribution of the Relative Concentrations of Bacteria',
    caption = 'Figure 2: Distribution of the relative concentrations of bacteria in Austin, TX'
  ) + 
  theme_bw()

  

water_tidy |> #All units are colonies/100 mL
  group_by(parameter) |>
  summarize(mean_concentration = mean(result),
            median_concentration = median(result),
            sd_concentration = sd(result),
            IQR_concentration = IQR(result))
```

Figure 2 and the summary statistics show us large variations in the
dangerous concentrations of bacteria in water sources in Austin. Median
concentrations of enterococcus are above the safe limit at 15,100
colonies/100 mL. Mean concentrations of total coliform bacteria and
fecal streptococci tell a similar story at 1,299.784 colonies/100 mL and
6,072.419 colonies/100 mL. However, the standard deviation for each
bacteria concentration is on the order of thousands and tens of
thousands, implying that there are still many locations of dangerously
high bacteria concentration.

Finally, we can look at the changing concentrations of bacteria over
time in both surface and ground water sources.

```{r}
water_tidy |>
  filter(medium == 'Ground Water' | medium == 'Surface Water' & medium != 'Storm Water') |>
  ggplot() + #Make two plots for each water source of results as they change per year
  geom_point(aes(x = year, y = result, color = medium), show.legend = F) +
  facet_wrap(~medium) +
  labs(
    x = 'Year',
    y = 'Bacteria Concentration (colonies/100 mL)',
    title = 'Bacteria Concentration Since 1991 Across Water Sources',
    caption = 'Figure 3: Bacteria concentration since 1991 across water sources'
  ) +
  theme_bw()
  
water_tidy |> #Surface water averages
  filter(medium == 'Surface Water') |>
  group_by(year) |>
  summarize(mean_concentration = mean(result),
            median_concentration = median(result),
            sd_concentration = sd(result),
            IQR_concentration = IQR(result))

water_tidy |> #Mean concentrations in colonies/100 mL, ground water averages
  filter(medium == 'Ground Water') |>
  group_by(year) |>
  summarize(mean_concentration = mean(result),
            median_concentration = median(result),
            sd_concentration = sd(result),
            IQR_concentration = IQR(result))
```

Surface water concentrations are much more frequently reported than
groundwater concentrations. Groundwater concentrations have stayed
approximately the same since the 20th century. Surface water
concentrations were extremely high in the 20th century (an average of
23,713.57 colonies/100 mL in 1994) and have consistently improved in the
21st century (an average of 212.04 colonies/100 mL in 2010). However,
there appears to be a slight uptick in concentration in recent years as
shown by the right tail in Figure 3 of surface water concentrations.

## Time of Day Bacteria/Pathogen Samples Are Taken

In order to see what time of day bacteria samples are taken at, we must
first look at the distribution of the time periods the data was
acquired. We will find the number of samples collected per year to see
this distribution.

```{r}
water_tidy |>
  ggplot() +
  geom_histogram(aes(x = year), binwidth = 2, center = 1, fill = 'lightblue') +
  scale_x_continuous(breaks = seq(1986, 2025, 3)) + #Show variations for every three years
  labs(
    x = 'Year',
    y = 'Frequency of Sample Collection',
    title = 'Distribution of Samples Collected over Time',
    caption = 'Figure 4: Distribution of samples collected over time'
  ) +
  theme_bw()

water_tidy |> #Count number of samples collected per year
  group_by(year) |>
  summarize(count = n(),
            percent = 100*n()/nrow(water_tidy))
```

We see an approximately unimodal distribution where more samples were
taken in the 90's and decreased in the 2000's. The most samples were
taken in 2009 at 1,857 samples. The least samples were taken in 1987 at
10 samples.

Next, we look at the distribution of samples taken per bacteria type. We
can make a barplot for the distribution and compare across the bars.

```{r}
water_tidy |>
  group_by(parameter) |>
  summarize(count = n()) |>
  ggplot() + #Bar plot showing number of samples that had a given bacteria concentration
  geom_bar(aes(x = count, y = parameter, fill = parameter), stat = 'identity', show.legend = F) +
  labs(
    x = 'Number of Samples',
    y = 'Type of Bacteria',
    title = 'Distribution of Bacteria Samples in Austin',
    caption = 'Figure 5: Distribution of bacteria samples in Austin, TX'
  ) +
  theme_bw()

water_tidy |>
  group_by(parameter) |> #count bacteria and percent in total
  summarize(count = n(),
            percent = 100*n()/nrow(water_tidy))
```

The bacteria that was found the most is E coli, in 14,852 samples, while
the least is enterococcus in only 167 samples. Next, we can find the
hour at which the samples were taken for each type of bacteria.

```{r}
water_tidy |>
  ggplot() +
  geom_bar(aes(x = hour(date), fill = parameter), show.legend = F) +
  scale_x_continuous(breaks = seq(0, 24, 2)) + # Change x axis to show more variation every 2 hours
  facet_wrap(~parameter) +
  theme_bw() +
  labs(
    x = 'Hour Since Midnight',
    y = 'Number of Samples',
    title = 'Distribution of Samples per Day Across Bacteria Type',
    caption = 'Figure 6: Distribution of samples per day across bacteria type'
  )

water_tidy |> #Find number of samples taken per hour and percent
  group_by(hour(date)) |>
  summarize(count = n(),
            percent = 100*n()/nrow(water_tidy))
```

For fecal streptococci, fecal coliform bacteria, and E coli, midnight is
the most frequent time for sample collection (15,988 samples). However,
this is likely the default time entered for when the exact collection
time is unknown. After the midnight peak, the next most frequent time is
around 11 AM and noon (3,165 samples).

## Bacteria/Pathogen Concentrations by Location

First, we will examine the distribution of bacteria/pathogen
concentrations across Austin, Texas.

```{r}
# Exploring all bacteria/pathogen concentrations in relation to location of samples
water_tidy |>
  #Plot lat/long map
  ggplot() +
  geom_point(aes(x = lon_dd_wgs84, y = lat_dd_wgs84)) +
  labs(
    x = 'Longitude (°)',
    y = 'Latitude (°)',
    title = 'Distribution of all Bacteria/Pathogen Concentrations across Austin',
    caption = 'Figure 7: Distribution of All Bacteria/Pathogen Concentrations across Austin'
  ) +
  theme_bw()

# Exploring all bacteria/pathogen concentrations in relation to location of samples by parameter
water_tidy |>
  # Filter out coordinates that are not in Austin
  filter(lon_dd_wgs84 != 0 | lat_dd_wgs84 != 0) |>
  #Plot lat/long map
  ggplot() +
  geom_point(aes(x = lon_dd_wgs84, y = lat_dd_wgs84, color = parameter), show.legend = FALSE) +
  labs(
    x = 'Longitude (°)',
    y = 'Latitude (°)',
    title = 'Distribution of Bacteria/Pathogen Concentrations across Austin by Parameter',
    caption = 'Figure 8: Distribution of Bacteria/Pathogen Concentrations across Austin by Parameter'
  ) +
  facet_wrap(~parameter) +
  theme_bw()
```

From Figures 7 and 8, there are higher bacteria/pathogen concentrations
in the central areas of Austin compared to the outer regions. This
pattern suggests several potential factors. The central part of Austin
contains a higher population density and increased human activity, which
can contribute to the proliferation of bacteria and pathogens.
Additionally, urban environments often have more sources of pollution
and waste, which can create favorable conditions for microbial growth.
The outer areas, being less densely populated and more suburban or
rural, may have fewer such sources, leading to lower bacteria/pathogen
concentrations.

## Safe Bacteria/Pathogen Concentrations Levels by Site Type

The US EPA sets specific recommendations for safe bacteria/pathogen
concentrations in various water types to protect public health. These
standards help ensure that water used for various purposes, from
recreation to consumption, remains safe and free from harmful levels of
pathogens. We will focus on three water types: recreational, treatment,
and drinking. Furthermore, we will look at sample counts and median
results for three different groups of water types: recreational water,
domestic water supply, and drinking water.

Figure 9 is a distribution of bacteria/pathogen concentrations by site
types in Austin, Texas.

```{r}
# Identifying the different site types
water_tidy |>
  group_by(site_type) |>
  ggplot() +
  geom_bar(aes(y = site_type, fill = site_type), show.legend = F) +
  labs(
    x = 'Frequency of Site Type',
    y = 'Site Type',
    title = 'Distribution of Site Types',
    caption = 'Figure 9: Distribution of site types in Austin, TX'
  ) + 
  theme_bw()

water_tidy |>
  group_by(site_type) |>
  summarize(count = n())
```

### Recreational Water

For recreational water, we will look at site types lakes, springs, and
streams because these waters are typically considered for recreational
activities such as swimming, boating, and fishing. We will compare on E.
coli and enterococcus concentrations with fecal coliform bacteria
concentrations because the EPA emphasizes the former as better
indicators.

```{r}
# Frequency of recreational water site types
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream')) |>
  group_by(site_type) |>
  summarize(sample_count = n())
```

The EPA's current standards in recreational water are a geometric mean
of **no more than 126 colony-forming units (CFU) per 100 mL for E.
coli** and a geometric mean of **no more than 35 CFU per 100 mL for
enterococcus**. The EPA previously recommended a geometric mean of **no
more than 200 CFU per 100 mL for fecal coliform bacteria**. In general,
these thresholds are particularly significant for water used in
activities where direct contact with skin is common. Maintaining below
these amounts ensure that the water is safe for recreational use,
reducing the risk of infections and other health issues associated with
exposure to contaminated water.

Note: Colonies/per 100 mL is the same as CFU/100 mL.

Information on safe concentrations of E. coli and enterococcus in
recreational water can be found here:
<https://www.epa.gov/sites/default/files/2015-10/documents/rec-factsheet-2012.pdf>.
Information on safe concentrations of fecal coliform bacteria in
recreational waters can be found here:
<https://www.knowyourh2o.com/outdoor-4/fecal-coliform-bacteria-in-water>/

```{r}
# Comparing all three parameters in terms of safeness in recreational water
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream')) |>
  filter(parameter %in% c("E COLI BACTERIA", "ENTEROCOCCUS", "FECAL COLIFORM BACTERIA")) |>
  mutate(safeness = case_when(
    parameter == "E COLI BACTERIA" & result <= 126 ~ T,
    parameter == "E COLI BACTERIA" & result > 126 ~ F,
    parameter == "ENTEROCOCCUS" & result <= 35 ~ T,
    parameter == "ENTEROCOCCUS" & result > 35 ~ F,
    parameter == "FECAL COLIFORM BACTERIA" & result <= 200 ~ T,
    parameter == "FECAL COLIFORM BACTERIA" & result > 200 ~ F)) |>
  ggplot() +
  #Plot lat/long map and color based on safety
  geom_point(aes(x = lon_dd_wgs84, y = lat_dd_wgs84, color = safeness)) +
  scale_color_manual(values = c("red", "forestgreen")) +
  labs(
    x = 'Longitude (°)',
    y = 'Latitude (°)',
    title = 'Distribution Comparing Safeness Levels for Recreational Water',
    caption = 'Figure 10: Comparison of Safeness Levels for Recreational Water in Austin, TX'
  ) +
  facet_wrap(~parameter) +
  theme_bw()

# Summary statistics
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream')) |>
  filter(parameter %in% c("E COLI BACTERIA", "ENTEROCOCCUS", "FECAL COLIFORM BACTERIA")) |>
  mutate(safeness = case_when(
    parameter == "E COLI BACTERIA" & result <= 126 ~ T,
    parameter == "E COLI BACTERIA" & result > 126 ~ F,
    parameter == "ENTEROCOCCUS" & result <= 35 ~ T,
    parameter == "ENTEROCOCCUS" & result > 35 ~ F,
    parameter == "FECAL COLIFORM BACTERIA" & result <= 200 ~ T,
    parameter == "FECAL COLIFORM BACTERIA" & result > 200 ~ F)) |>
  group_by(parameter, safeness) |>
  summarize(sample_count = n(), 
            median_result = median(result),
            iqr_result = IQR(result),
            .groups = 'drop') |>
  group_by(parameter) |>
  mutate(parameter_count = sum(sample_count),
         proportion = sample_count/parameter_count)
```

For E. coli bacteria, 71.1% of the samples (9771 out of 13739) are
considered safe for recreational use, and 28.9% of the samples (3968 out
of 13739) are considered unsafe for recreational use. For enterococcus,
only 5.3% of the samples (2 out of 38) are considered safe for
recreational use, and 94.7% of the samples (36 out of 38) are considered
unsafe for recreational use. For fecal coliform bacteria, 76.0% of the
samples (10194 out of 13408) are considered safe for recreational use,
and 24.0% of the samples (3214 out of 13408) are considered unsafe for
recreational use. Therefore, the majority of samples for E. coli and
fecal coliform bacteria are within safe levels for recreational water
sites. A significant portion of enterococcus samples are above safe
levels, indicating potential health risks for recreational activities in
those waters. However, it is difficult to analyze enterococcus samples
because there are only 38 total samples.

Safe E. coli bacteria samples (the majority) have a median result of
32.5177 CFU/100 mL, which is significantly less than 126 CFU/100 mL,
suggesting that most of the samples are within the safeness threshold.
Safe fecal coliform bacteria samples (the majority) have a median result
44.9287 CFU/100 mL, which is significantly less than 200 CFU/100 mL,
suggesting that most of the samples are within the safeness threshold.
However, the median result for unsafe E. coli bacteria samples is 387
CFU/100 mL, and the median result for unsafe fecal coliform bacteria
samples is 620 CFU/100 mL. These values are significantly higher than
the safeness threshold. Entercoccus median results are disgarded because
there are not enough samples to conduct a proper analysis.

### Domestic Water Supply

Domestic water supply refers to the provision of water for household
uses such as drinking, cooking, cleaning, and bathing. This water
typically comes from various sources, including wells, springs, lakes,
and streams. Before it reaches homes, the water undergoes rigorous
treatment processes to ensure it meets safety standards. These processes
include filtration, sedimentation, and disinfection to remove
contaminants and pathogens. Ensuring a reliable and safe domestic water
supply is crucial for public health and well-being, as it prevents
waterborne diseases and supports daily life activities. Proper
management and treatment of water sources are essential to maintain the
quality and availability of this vital resource.

For domestic water supply, we will look at site types lakes, springs,
streams, and wells. The EPA's current standard in
domestic water supply is **fewer than 2000 colonies/100 mL**.

```{r}
# Frequency of domestic water supply site types
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream', 'Well')) |>
  group_by(site_type) |>
  summarize(sample_count = n())
```

Information on safe concentrations of fecal coliform bacteria in
recreational waters can be found here:
<https://www.knowyourh2o.com/outdoor-4/fecal-coliform-bacteria-in-water>/

```{r}
# Comparing all fecal coliform bacteria concentration levels for domestic water supply
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream', 'Well')) |>
  filter(parameter == "FECAL COLIFORM BACTERIA") |>
  mutate(safeness = ifelse(result < 2000, T, F)) |>
  ggplot() +
  #Plot lat/long map and color based on safety
  geom_point(aes(x = lon_dd_wgs84, y = lat_dd_wgs84, color = safeness)) +
  scale_color_manual(values = c("red", "forestgreen")) +
  labs(
    x = 'Longitude (°)',
    y = 'Latitude (°)',
    title = 'Distribution of Fecal Coliform Bacteria Safeness Levels in Domestic Water Supply',
    caption = 'Figure 11: Distribution of Fecal Coliform Bacteria Safeness Levels in Domestic Water Supply in Austin, TX'
  ) +
  facet_wrap(~parameter) +
  theme_bw()

# Summary statistics
water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream', 'Well')) |>
  filter(parameter == "FECAL COLIFORM BACTERIA") |>
  mutate(safeness = ifelse(result < 2000, T, F)) |>
  group_by(site_type, safeness) |>
  summarize(sample_count = n(), 
            median_result = median(result),
            iqr_result = IQR(result),
            .groups = 'drop') |>
  group_by(site_type) |>
  mutate(site_type_count = sum(sample_count),
         proportion = sample_count/site_type_count)
```

For lakes, 95.4% of the samples (1092 out of 1145) are considered safe
for domestic water supply, and 4.6% of the samples (53 out of 1145) are
considered unsafe for domestic water supply. For springs, 99.9% of the
samples (3469 out of 3472) are considered safe for domestic water
supply, 0.1% of the samples (3 out of 3472) are considered unsafe for
domestic water supply. For streams, 91.7% of the samples (8046 out of
8771) are considered safe for domestic water supply, 8.3% of the samples
(725 out of 8771) are considered unsafe for domestic water supply. For
wells, 98.0% of the samples (50 out of 51) are considered safe for
domestic water supply, 2.0% of the samples (1 out of 51) are considered
unsafe for domestic water supply. Therefore, the majority of samples
from lakes, springs, streams, and wells are within safe levels for
domestic water supply. This indicates that these sources are generally
reliable for providing safe water for household use after appropriate
treatment.

The median safe result values for all site types, which consists of the
majority of samples for each site type, are significantly fewer than
2000 CFU/100 mL. The median result for lakes is 60 CFU/100 mL, the
median result for springs is 13 CFU/100 mL, the median result for
streams is 59 CFU/100 mL, and the median result for wells is 2 CFU/100
mL. The median unsafe result for wells, which is 2100 CFU/100 mL, is
slightly above the safeness threshold and is the lowest of the four
water site, but this is based only on one sample. This suggests the need
for continuous monitoring and rigorous treatment processes to ensure
safety. Streams have the highest median result of 6820 CFU/100 mL among
unsafe samples, suggesting they are the least suitable option for
domestic water supply.

### Drinking Water

For drinking water, we will look at site types drinking water tap and
well because these are typically considered direct sources of drinking
water. The EPA's current standard in drinking water is the **absence of
E. coli**.

Information on safe concentrations of E. coli bacteria in recreational
waters can be found here:
<https://www.knowyourh2o.com/outdoor-4/fecal-coliform-bacteria-in-water>/

```{r}
# Frequency of drinking water site types
water_tidy |>
  filter(parameter == 'E COLI BACTERIA') |>
  filter(site_type %in% c('Drinking Water Tap', 'Well')) |>
  group_by(site_type) |>
  summarize(count = n())
```

```{r}
# Comparing all E. coli concentration levels in drinking water
water_tidy |>
  filter(parameter == 'E COLI BACTERIA') |>
  filter(site_type %in% c('Drinking Water Tap', 'Well')) |>
  mutate(safeness = ifelse(result == 0, T, F)) |>
  ggplot() +
  #Plot lat/long map and color based on safety
  geom_point(aes(x = lon_dd_wgs84, y = lat_dd_wgs84, color = safeness)) +
  scale_color_manual(values = c("red", "forestgreen")) +
  labs(
    x = 'Longitude (°)',
    y = 'Latitude (°)',
    title = 'Distribution of E. Coli Safeness Levels for Drinking Water ',
    caption = 'Figure 12: Distribution of E. Coli Safeness Levels for Drinking Water in Austin, TX'
  ) +
  theme_bw()

# Summary statistics
water_tidy |>
  filter(parameter == 'E COLI BACTERIA') |>
  filter(site_type %in% c('Drinking Water Tap', 'Well')) |>
  mutate(safeness = ifelse(result == 0, T, F)) |>
  group_by(site_type, safeness) |>
  summarize(sample_count = n(), 
            median_result = median(result),
            iqr_result = IQR(result),
            .groups = 'drop')
```

All drinking water sites have unsafe E. coli concentrations (greater
than 0 CFU/100 mL or 0 MPN/100ML). Therefore, it would be hazardous to
drink from these direct sources of water.

------------------------------------------------------------------------

# Modeling (Nabil)

```{r}
#only looking at ground vs. surface water
#Also only filter down to site types that represent a large portion of the data, remove any single row observations
water_tidy <- water_tidy |>
  filter(medium == 'Ground Water' | medium == 'Surface Water' & medium != 'Storm Water') |>
  filter(site_type == 'Well' | site_type == 'Urban runoff' | site_type == 'Stream' | site_type == 'Spring' | site_type == 'Lake' & site_type != 'BMP - Wet Pond')
## Decision Tree Model

#Make decision tree with result as outcome and sample_date, medium, and site_type as predictors
tree_reg <- rpart(result ~ date + medium + site_type,
                    data = water_tidy)

#Visualize decision tree
rpart.plot(tree_reg)

#Calculate RMSE
sqrt(mean((water_tidy$result - predict(tree_reg, water_tidy))^2))

# Initialize a vector to keep track of the performance for each k-fold
perf_k <- NULL

## 5-fold Cross-Validation

# Choose number of folds
k = 5 

# To have the same random sample, use set.seed
set.seed(322)

# Randomly order rows in the dataset
shuffled_data <- water_tidy[sample(nrow(water_tidy)), ] 

# Create k folds in the dataset
shuffled_data <- water_tidy |>
  mutate(folds = cut(seq(1:nrow(shuffled_data)), breaks = k, labels = FALSE))

# Use a for-loop to get performance for each k-fold
for(i in 1:k){
  # Split data into train and test data
  train_i <- shuffled_data |> filter(folds != i) # train data = all observations except in fold i
  test_i <- shuffled_data |> filter(folds == i)  # test data = observations in fold i
  
  # Train model on train data (all but fold i)
  train_model <- rpart(result ~ date + medium + site_type,
                    data = train_i)
  
  # Performance listed for each test data = fold i
  # Save the RMSE values
  perf_k[i] <- sqrt(mean((water_tidy$result - predict(train_model, test_i))^2))
}

# Performance for each fold 
perf_k

# Average performance over all k folds
mean(perf_k)
sd(perf_k)
```

# Modeling (Amanda)

```{r}
# Focus on recreational waters
rec_water_tidy <- water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream')) |>
  filter(parameter %in% c("E COLI BACTERIA", "ENTEROCOCCUS", "FECAL COLIFORM BACTERIA")) |>
  mutate(safeness = case_when(
    parameter == "E COLI BACTERIA" & result <= 126 ~ 1,
    parameter == "E COLI BACTERIA" & result > 126 ~ 0,
    parameter == "ENTEROCOCCUS" & result <= 35 ~ 1,
    parameter == "ENTEROCOCCUS" & result > 35 ~ 0,
    parameter == "FECAL COLIFORM BACTERIA" & result <= 200 ~ 1,
    parameter == "FECAL COLIFORM BACTERIA" & result > 200 ~ 0))

# Decision tree model for recreational waters
fit_tree_rec <- rpart(safeness ~ result + site_type + parameter,
                    data = rec_water_tidy,
                    method = 'class')

# ROC curve
ROC <- ggplot(rec_water_tidy) + 
    geom_roc(aes(
      # Outcome is safeness
      d = safeness,
      # Probability of safeness based on the decision tree model
      m = predict(fit_tree_rec, rec_water_tidy)[ ,2]))
ROC

# Visualize the decision tree
rpart.plot(fit_tree_rec)

# Calculate the area under the curve
calc_auc(ROC)$AUC

# Choose number of folds
k = 5 

# To have the same random sample, use set.seed
set.seed(322)

# Randomly order rows in the dataset
shuffled_data <- rec_water_tidy[sample(nrow(rec_water_tidy)), ] 

# Create k folds in the dataset
shuffled_data <- rec_water_tidy |>
  mutate(folds = cut(seq(1:nrow(shuffled_data)), breaks = k, labels = FALSE))

# Initialize a vector to keep track of the performance for each k-fold
perf_k <- NULL

# Use a for-loop to get performance for each k-fold
for(i in 1:k){
  # Split data into train and test data
  train_i <- shuffled_data |> filter(folds != i) # train data = all observations except in fold i
  test_i <- shuffled_data |> filter(folds == i)  # test data = observations in fold i
  
  # Train model on train data (all but fold i)
  # CHANGE: what model/predictors should be included
  train_model <- rpart(safeness ~ result + site_type + parameter,
                    data = train_i,
                    method = 'class')
  
  # Performance listed for each test data = fold i
  # CHANGE: how the performance is calculated
  ROC <- ggplot(test_i) + 
    geom_roc(aes(
      # Outcome is safeness
      d = safeness,
      # Probability of safeness based on the decision tree model
      m = predict(train_model, test_i)[ ,2]))
  # Save the AUC values
  perf_k[i] <- calc_auc(ROC)$AUC
}

# Performance for each fold 
perf_k

# Average performance over all k folds
mean(perf_k)
sd(perf_k)
```

```{r}
# Focus on domestic water supply
domestic_water_tidy <- water_tidy |>
  filter(site_type %in% c('Lake', 'Spring', 'Stream', 'Well')) |>
  filter(parameter == "FECAL COLIFORM BACTERIA") |>
  mutate(safeness = ifelse(result < 2000, 1, 0))

# Decision tree model for domestic water supply
fit_tree_domestic <- rpart(safeness ~ result + site_type + parameter,
                    data = domestic_water_tidy,
                    method = 'class')

# ROC curve
ROC <- ggplot(domestic_water_tidy) + 
    geom_roc(aes(
      # Outcome is safeness
      d = safeness,
      # Probability of safeness based on the decision tree model
      m = predict(fit_tree_domestic, domestic_water_tidy)[ ,2]))
ROC

# Visualize the decision tree
rpart.plot(fit_tree_domestic)

# Calculate the area under the curve
calc_auc(ROC)$AUC
# Choose number of folds
k = 5 

# To have the same random sample, use set.seed
set.seed(322)

# Randomly order rows in the dataset
shuffled_data <- domestic_water_tidy[sample(nrow(domestic_water_tidy)), ] 

# Create k folds in the dataset
shuffled_data <- domestic_water_tidy |>
  mutate(folds = cut(seq(1:nrow(shuffled_data)), breaks = k, labels = FALSE))

# Initialize a vector to keep track of the performance for each k-fold
perf_k <- NULL

# Use a for-loop to get performance for each k-fold
for(i in 1:k){
  # Split data into train and test data
  train_i <- shuffled_data |> filter(folds != i) # train data = all observations except in fold i
  test_i <- shuffled_data |> filter(folds == i)  # test data = observations in fold i
  
  # Train model on train data (all but fold i)
  # CHANGE: what model/predictors should be included
  train_model <- rpart(safeness ~ result + site_type + parameter,
                    data = train_i,
                    method = 'class')
  
  # Performance listed for each test data = fold i
  # CHANGE: how the performance is calculated
  ROC <- ggplot(test_i) + 
    geom_roc(aes(
      # Outcome is safeness
      d = safeness,
      # Probability of safeness based on the decision tree model
      m = predict(train_model, test_i)[ ,2]))
  # Save the AUC values
  perf_k[i] <- calc_auc(ROC)$AUC
}

# Performance for each fold 
perf_k

# Average performance over all k folds
mean(perf_k)
sd(perf_k)
```
------------------------------------------------------------------------

# Discussion

After conducting this analysis, we have a better picture of
bacteria/pathogen concentrations in Austin. We can now see the changes
across day and water source. To begin with, Figure 1 tells us that the
majority of water sources are surface water sources (22,531 sources).
This tells us that our focus should be on testing surface water sources
since they affect the most people. This is supported by Figure 3 where
we see that almost all sampling has historically been done on surface
water sources. However, Figures 3 and 4 also indicate that testing
efforts have waned in the 21st century. While this is likely partly due
to improved experimental analysis methods, it still represents a danger
to the community since contamination could be missed with less testing.

To see the variations in a day, we can look to Figures 5 and 6. E coli
bacteria are highly present in the samples, representing 14,852 samples.
Samples are most often taken in the late morning, between 11 AM and noon
(ignoring the midnight category). The trend makes sense given that these
times are during the workday and when most people are active. However,
it is important to note that this is the warmest part of the day.
Sampling methods will be affected by the temperature and properties of
the water. Further research needs to be done on how this trend affects
the measured bacteria concentration. The trends approximately agreed
with expected findings of most frequent sampling happening during the
day. We also did expect highest bacteria concentrations to appear in
water sources closest to the outside environment. This was proven by the
data and the dominance of surface water sources.

We can draw several important conclusions regarding the safeness levels
of various water sources for both recreational and domestic use from
Figures 10-12. For recreational water sites, the majority of samples for
E. coli and fecal coliform bacteria are within safe levels, with 71.1%
and 76.0% of samples respectively being safe. Enterococcus contamination
poses a significant risk, with only 5.3% of samples being safe. However,
there are not enough enterococcus samples (only 38) to make a reasonable
conclusion. Nonetheless, this highlights the need for improved water
quality measures in recreational areas.

For domestic water supply, most sources such as lakes, springs, streams,
and wells are generally safe, with high percentages of safe samples
ranging from 91.7% to 99.9%. However, occasional instances of unsafe
samples from drinking water taps and wells underscore the importance of
continuous monitoring and rigorous treatment processes to ensure the
safety of domestic water supplies. Ensuring safe water for both
recreational and domestic use is vital for preventing waterborne
diseases and maintaining overall public health. Continuous efforts in
monitoring, treatment, and management of water sources are essential to
achieve this goal.

Therefore, the data supports the hypothesis that water sites exposed to
urban influences, such as recreational water, domestic water supply, and
drinking water, mainly have safe bacteria/pathogen concentration levels.
However, there are some samples with high median unsafe results
compared. y, and drinking water), which is more exposed to urban
influences, contains some high median unsafe results. For instance,
streams, which are often impacted by urban runoff and other sources of
contamination, have a high unsafe median result of 6820 CFU/100 mL for
domestic water supply. This indicates that streams in urban areas are
particularly vulnerable to contamination and may require more rigorous
monitoring and treatment to ensure safety.

------------------------------------------------------------------------

# Reflection, Acknowledgements, and References

## Reflection

From the process, we have learned the importance of meticulous data
analysis and the need for clear, concise communication. Interpreting
data accurately is essential to draw meaningful conclusions and provide
actionable insights. Moreover, this project underscored the significance
of continuous monitoring and rigorous treatment processes to ensure
water safety. It also highlighted the critical role of adhering to
regulatory standards to protect public health.

The exploratory data analysis also revealed the amount of options
available to tell the story of the data. Different visualizations offer
a unique perspective on the dataset, giving us endless options to work
with. However, we also learned that data is never perfect nor is it
recorded perfectly. A large part of the work went into wrangling the
dataset.

Overall, this project has reinforced the value of detailed analysis,
effective data interpretation, and the importance of maintaining high
standards in water quality management.

### Challenges

The most challenging aspect of this project was ensuring the accurate
interpretation and integration of the data set. Understanding the
specific safety standards and guidelines for different types of water
required a thorough grasp of regulations/standards and their practical
implications.

Additionally, wrangling the data presented an initial hurdle to our
analysis. The original dataset contained too many rows and needed to be
filtered. Even after filtering, the dataset was not tidy and needed to
have columns renamed, rows dropped, and new columns made.

### Contributions

Amanda:

-   Method

-   Individual research question

-   Individual results

-   Individual discussion

-   Individual reflection

-   Acknowledgments

-   References

Nabil:

-   Introduction

-   Method

-   Individual research question

-   Individual results

-   Individual discussion

-   Individual reflection

-   Acknowledgments

## Acknowledgements

We would like to extend our sincere gratitude to the City of Austin for
providing access to their Open Data Portal. This invaluable resource has
been instrumental in our analysis, allowing us to access comprehensive
and up-to-date data on water quality and other critical environmental
metrics. The transparency and accessibility of this data empower
researchers, policymakers, and the community to make informed decisions
and take proactive measures to ensure public health and safety. Thank
you, City of Austin, for your commitment to open data and for supporting
initiatives that enhance the well-being of our community.

## References

### City of Austin Open Data Portal:

-   <https://data.austintexas.gov/Environment/Water-Quality-Sampling-Data/5tye-7ray/about_data>

### Background Context:

-   <https://www.epa.gov/sites/default/files/2015-10/documents/rec-factsheet-2012.pdf>

-   <https://www.knowyourh2o.com/outdoor-4/fecal-coliform-bacteria-in-water>

-   <https://torrentlab.com/what-is-the-purpose-of-stormwater-testing-and-when-is-it-required/>

### Image:

-   <https://data.austintexas.gov/Environment/Water-Quality-Sampling-Data/5tye-7ray/about_data>

------------------------------------------------------------------------

# Formatting

-   Create the report using R Markdown, with headers for each section,
    and comments of the R code.

-   The final report should be at least 5 pages but no more than 20
    pages (the number of pages can vary greatly depending on the
    cleaning process though). It is extremely important that you select
    pages to corresponding sections when submitting on Gradescope.

-   If a team project, identify all team members on the submission.
